Extracting production locations for each species to obtain environmental forcings

### 2025-07-10 02_extracting_production_locations.qmd
Consolidating

```{r setup}
library(tidyr)
library(dplyr)
library(magrittr)
library(here)
library(janitor)
library(sf)
library(qs)
library(countrycode)
library(terra)
library(units)
library(readr)

@@

Clean aquaculture production data from the FAO and isolate production data for Atlantic salmon.

@@

quantity <- quantity %>% 
  merge(environment_codes, by.x = "environment_alpha_2_code", by.y = "Code") %>%
  rename(environment = Name_En) %>% 
  filter(environment == "Marine") %>% 
  select(-environment_alpha_2_code)

# Apply species codes and filter by relevant species groups
species_codes <- file.path(rawdata_path, "Aquaculture_2025.1.0", "CL_FI_SPECIES_GROUPS.csv") %>% 
  read.csv() %>% 
  filter(str_detect(Scientific_Name, "Salmo")) %>% 
  filter(Scientific_Name %in% c("Salmoniformes (=Salmonoidei)", "Salmo spp", "Salmo salar", "Salmonidae")) %>% 
  distinct(X3A_Code, Scientific_Name)

@@

# Apply status codes and filter by relevant species groups
status_codes <- file.path(rawdata_path, "Aquaculture_2025.1.0", "CL_FI_SYMBOL_SDMX.csv") %>% 
  read.csv() %>% 
  distinct(Symbol, Name_En, Description_En)

quantity <- quantity %>% 
  merge(status_codes, by.x = "status", by.y = "Symbol") %>% 
  rename(status_nm = Name_En,
         status_description = Description_En) %>% 
  select(-status)
```

Pull in farm locations, add FAO Major fishing regions

```{r}
farms_sf <- qread(file = "data/_general_data/farm_locations/locations.qs") %>% 
  st_as_sf(coords = c("X", "Y"), crs = "EPSG:4326") %>% 
  mutate(row_num = row_number()) %>% 
  group_by(row_num) %>% 
  group_split()

@@

# filter country level production for all species in non-inland waters where production is greater than 0 tonnes.

mariculture_prod_all <- 
    aquaculture_raw %>% 
    clean_names() %>% 
    select(c(1:7, starts_with("x"))) %>% 
    pivot_longer(names_to = "year", values_to = "production", cols= -c(1:7)) %>% 
    mutate(year = gsub(pattern="x", replacement = "", year)) %>% 
    filter(year %in% c(2017)) %>% 
    filter(!grepl("Inland waters", fao_major_fishing_area_name)  & 
             !asfis_species_name_3 %in% c("Clams, cockles, arkshells", 
                                          "Abalones, winkles, conchs",
                                          "Mussels", 
                                          "Oysters",
                                          "Shrimps, prawns",
                                          "Crabs, sea-spiders",
                                          "Freshwater molluscs",
                                          "Squids, cuttlefishes, octopuses",
                                          "Pearls, mother-of-pearl, shells",
                                          "Miscellaneous aquatic plants",
                                          "Scallops, pectens"),
           !grepl("seaweeds" , asfis_species_name_3),
           !grepl("molluscs|cucumber|invertebrates", asfis_species_name_2)) %>% 
    group_by(country_name, asfis_species_name_2, asfis_species_name_3, fao_major_fishing_area_name, fao_major_fishing_area_code, year) %>% 
    summarise(production = sum(production)) %>% 
    ungroup() %>% 
    mutate(iso3c = countrycode(sourcevar = country_name, origin = "country.name", destination = "iso3c", warn =TRUE)) %>% 
    group_by(iso3c, country_name, year) %>% 
    nest() %>% 
    mutate(national_production = map(data, ~(sum(.$production)))) %>% 
    unnest(cols = c(data, national_production)) %>% 
    ungroup() %>% 
  filter(production>0) %>% 
  filter(!asfis_species_name_2 %in% c("Marine fishes nei")) %>% 
   mutate(model_name = case_when(grepl("rouper", asfis_species_name_2) ~ "general_grouper", 
                                asfis_species_name_2 %in% c("Coho(=Silver) salmon", "Rainbow trout") ~ "general_salmonid",
                                TRUE ~ gsub(pattern = " ", replacement = "_", tolower(asfis_species_name_2)))) %>% 
  relocate(model_name, .before = asfis_species_name_2) %>% 
  mutate(farm_name = case_when(model_name %in% c("atlantic_salmon", "general_salmonid") ~ "Salmonidae fish",
                                 TRUE ~ "General marine fish")) %>% 
 filter(asfis_species_name_2 %in% spp_names) %>% 
  mutate(prop_production = production/national_production)

#Check that farm locations in mapped data are in the same FAO fishing regions as the production data

@@

>>>>

```{r setup}
library(tidyr)
library(dplyr)
library(magrittr)
library(here)
library(janitor)
library(sf)
library(qs)
library(countrycode)
library(terra)
library(units)
library(readr)
library(ggplot2)

@@

# Get production data

Clean aquaculture production data from the FAO and isolate data for Atlantic salmon.

@@

quantity <- quantity %>% 
  merge(environment_codes, by.x = "environment_alpha_2_code", by.y = "Code") %>%
  rename(environment = Name_En) %>% 
  select(-environment_alpha_2_code)

# Apply species codes
species_codes <- file.path(rawdata_path, "Aquaculture_2025.1.0", "CL_FI_SPECIES_GROUPS.csv") %>% 
  read.csv() %>% 
  distinct(X3A_Code, Scientific_Name)

@@

# Apply status codes
status_codes <- file.path(rawdata_path, "Aquaculture_2025.1.0", "CL_FI_SYMBOL_SDMX.csv") %>% 
  read.csv() %>% 
  distinct(Symbol, Name_En, Description_En)

quantity <- quantity %>% 
  merge(status_codes, by.x = "status", by.y = "Symbol") %>% 
  rename(status_nm = Name_En,
         status_description = Description_En) %>% 
  select(-status)

# Apply country and area codes
country_codes <- file.path(rawdata_path, "Aquaculture_2025.1.0", "CL_FI_COUNTRY_GROUPS.csv") %>% 
  read.csv() %>% 
  select(UN_Code, Identifier, Name_En, ISO3_Code)
area_codes <- file.path(rawdata_path, "Aquaculture_2025.1.0", "CL_FI_WATERAREA_GROUPS.csv") %>% 
  read.csv() %>% 
  select(Code, Name_En)

quantity <- quantity %>% 
  merge(country_codes, by.x = "country_un_code", by.y = "UN_Code", all.x = T, all.y = F) %>% 
  merge(area_codes, by.x = "area_code", by.y = "Code", all.x = T, all.y = F) %>% 
  rename(
    country = Name_En.x, 
    fao_fishing_area = Name_En.y,
    country_id = Identifier
    )

quantity <- quantity %>% 
  mutate(
    environment = as.factor(environment),
    Scientific_Name = as.factor(Scientific_Name),
    status_nm = as.factor(status_nm),
    status_description = as.factor(status_description),
    country = as.factor(country),
    fao_fishing_area = as.factor(fao_fishing_area),
    ISO3_Code = as.factor(ISO3_Code)
  ) %>% 
  filter(
    Scientific_Name %in% c("Salmoniformes (=Salmonoidei)", "Salmo spp", "Salmo salar", "Salmonidae") &
    environment == "Marine" & 
    status_nm != "Not significant (<0.5)"
    )

qsave(quantity, file.path(species_path, "production", "aquaculture_quantity.qs"))
```

Average production over years 2019:2023.

```{r aquaculture-production}
prod <- quantity %>% 
  filter(between(period, 2019, 2023)) %>% 
  group_by(country, Scientific_Name, environment, ISO3_Code, fao_fishing_area) %>% 
  reframe(production = mean(value))

qsave(prod, file.path(species_path, "production", "aquaculture_production.qs"))

ggplot(prod, aes(x = country, y = production)) +
  geom_point() +
  coord_flip()
```

# Match farm locations to FAO regions

Pull in farm locations, add FAO Major fishing regions.

```{r FAO-fishing-regions-of-farms}
farms_sf <- qread(file = "data/_general_data/farm_locations/locations.qs") %>% 
  st_as_sf(coords = c("X", "Y"), crs = "EPSG:4326") %>% 
  mutate(row_num = row_number()) %>% 
  group_by(row_num) %>% 
  group_split()

@@

```{r fix-misalignment}
# filter country level production for all species in non-inland waters where production is greater than 0 tonnes.
mariculture_prod_all <- aquaculture_raw %>% 
  clean_names() %>% 
  select(c(1:7, starts_with("x"))) %>% 
  pivot_longer(names_to = "year", values_to = "production", cols= -c(1:7)) %>% 
  mutate(year = gsub(pattern="x", replacement = "", year)) %>% 
  filter(year %in% c(2017)) %>% 
  filter(!grepl("Inland waters", fao_major_fishing_area_name)  & 
            !asfis_species_name_3 %in% c("Clams, cockles, arkshells", 
                                        "Abalones, winkles, conchs",
                                        "Mussels", 
                                        "Oysters",
                                        "Shrimps, prawns",
                                        "Crabs, sea-spiders",
                                        "Freshwater molluscs",
                                        "Squids, cuttlefishes, octopuses",
                                        "Pearls, mother-of-pearl, shells",
                                        "Miscellaneous aquatic plants",
                                        "Scallops, pectens"),
          !grepl("seaweeds" , asfis_species_name_3),
          !grepl("molluscs|cucumber|invertebrates", asfis_species_name_2)) %>% 
  group_by(country_name, asfis_species_name_2, asfis_species_name_3, fao_major_fishing_area_name, fao_major_fishing_area_code, year) %>% 
  summarise(production = sum(production)) %>% 
  ungroup() %>% 
  mutate(iso3c = countrycode(sourcevar = country_name, origin = "country.name", destination = "iso3c", warn =TRUE)) %>% 
  group_by(iso3c, country_name, year) %>% 
  nest() %>% 
  mutate(national_production = map(data, ~(sum(.$production)))) %>% 
  unnest(cols = c(data, national_production)) %>% 
  ungroup() %>% 
  filter(production>0) %>% 
  filter(!asfis_species_name_2 %in% c("Marine fishes nei")) %>% 
  mutate(model_name = case_when(grepl("rouper", asfis_species_name_2) ~ "general_grouper", 
                              asfis_species_name_2 %in% c("Coho(=Silver) salmon", "Rainbow trout") ~ "general_salmonid",
                              TRUE ~ gsub(pattern = " ", replacement = "_", tolower(asfis_species_name_2)))) %>% 
  relocate(model_name, .before = asfis_species_name_2) %>% 
  mutate(farm_name = case_when(model_name %in% c("atlantic_salmon", "general_salmonid") ~ "Salmonidae fish",
                                TRUE ~ "General marine fish")) %>% 
  filter(asfis_species_name_2 %in% spp_names) %>% 
  mutate(prop_production = production/national_production)

# Check that farm locations in mapped data are in the same FAO fishing regions as the production data

@@

# From Rich

```{r fix-misalignment}
# filter country level production for all species in non-inland waters where production is greater than 0 tonnes.
mariculture_prod_all <- aquaculture_raw %>% 
  clean_names() %>% 
  select(c(1:7, starts_with("x"))) %>% 
  pivot_longer(names_to = "year", values_to = "production", cols= -c(1:7)) %>% 
  mutate(year = gsub(pattern="x", replacement = "", year)) %>% 
  filter(year %in% c(2017)) %>% 
  filter(!grepl("Inland waters", fao_major_fishing_area_name)  & 
            !asfis_species_name_3 %in% c("Clams, cockles, arkshells", 
                                        "Abalones, winkles, conchs",
                                        "Mussels", 
                                        "Oysters",
                                        "Shrimps, prawns",
                                        "Crabs, sea-spiders",
                                        "Freshwater molluscs",
                                        "Squids, cuttlefishes, octopuses",
                                        "Pearls, mother-of-pearl, shells",
                                        "Miscellaneous aquatic plants",
                                        "Scallops, pectens"),
          !grepl("seaweeds" , asfis_species_name_3),
          !grepl("molluscs|cucumber|invertebrates", asfis_species_name_2)) %>% 
  group_by(country_name, asfis_species_name_2, asfis_species_name_3, fao_major_fishing_area_name, fao_major_fishing_area_code, year) %>% 
  summarise(production = sum(production)) %>% 
  ungroup() %>% 
  mutate(iso3c = countrycode(sourcevar = country_name, origin = "country.name", destination = "iso3c", warn =TRUE)) %>% 
  group_by(iso3c, country_name, year) %>% 
  nest() %>% 
  mutate(national_production = map(data, ~(sum(.$production)))) %>% 
  unnest(cols = c(data, national_production)) %>% 
  ungroup() %>% 
  filter(production>0) %>% 
  filter(!asfis_species_name_2 %in% c("Marine fishes nei")) %>% 
  mutate(model_name = case_when(grepl("rouper", asfis_species_name_2) ~ "general_grouper", 
                              asfis_species_name_2 %in% c("Coho(=Silver) salmon", "Rainbow trout") ~ "general_salmonid",
                              TRUE ~ gsub(pattern = " ", replacement = "_", tolower(asfis_species_name_2)))) %>% 
  relocate(model_name, .before = asfis_species_name_2) %>% 
  mutate(farm_name = case_when(model_name %in% c("atlantic_salmon", "general_salmonid") ~ "Salmonidae fish",
                                TRUE ~ "General marine fish")) %>% 
  filter(asfis_species_name_2 %in% spp_names) %>% 
  mutate(prop_production = production/national_production)

# Check that farm locations in mapped data are in the same FAO fishing regions as the production data
